Exploring United States Housing Affordability

### 2016-08-10 paper/Paper.Rmd
grammer and pushing

library(ggplot2)
library(dplyr)
library(readr)
library(stringr)
library(xml2)
library(rvest)
library(tidyr)
library(purrr)
library(scales)
library(maps)
library(ggthemes)
```

@@

  The dataset was taken from the American Housing Survey, which derives sets of data from 1985 and on. The data set includes categorizing each housing unit by factors such as affordability, income through Fair Market Rent and the Adjusted Median Income. Taking into account variables such as Burden (housing cost divided by its monthly income), certain households with a zero or negative income are input as -1 within the data. The data source was recorded from the American Housing Survey. Certain factors, such as Poverty Income, is based on the Census Bureau's official thresholds.

@@

#Abstract

@@

  The combined_years data was read into a dataframe using the read_csv function. Within the read_csv function, the column types were explicitly stated as a string variable (c = character, i = integer, etc.), in order for the data to be wrangled properly. Also, the nonsensical integers (negative numbers) were converted to NAs in the dataset so that the output data and graphs made stastical sense - negative numbers where the values should be positive skew the dataset. Then, the combined years dataset was cleaned using the piping methods in dplyr. The data was rather messy, with unecessary quotations around character values and inconsistent letter cases in the column names. This was rectified by setting all the column names to lowercase and removing all the quotations from character values. One the main premises of this paper is comparing the burden of household that are owning and renting. Since there was no explicit labeling of data corresponding to owning and renting households, an  additional column "ownrent" was created which labeled levels 1 and 2 to own and rent as factors. Finally, the data was written to a clean_years csv file.

@@

  The main metric that is referred to in this paper is burden. Burden is defined by the proportion of household income that is spent towards the cost of housing. This metric is then further dissected by analyzing against various variables and collating data on contributing factors, such as income. The first data explored is income data across the four designated regions in the United States (Northeast, Midwest, South, West). The adjusted incomes (per persons and per bedroom) are also considered, as certain regions may have different averages for household and property sizes, which could skew the unadjusted income data. The behavior of household income, bracketed in ranges in accordance to census data, over time is displayed to provide a better understanding of the changes in burden over time. The household incomes are then compared versus burden over time. here it is found that the ghiher income levels generally have lower housing burdens, and those burdens actually decrased over time, while burdens in the lower income brackets were generally higher and increased over time. Also, the housing burdens in the Northeast and West are generally higher than the Midwest and South. finally, to elucidate the gap in burden levels across high and low incomes, the housheold burden amongst owners and renters, and those with and without governement assistance is considered. Here it is found thatowning households tend to have lower burden levels, aso do those households with no assistance.

@@

  The inital plot measures the housing burden levels across the United States:

```{r, echo = F, fig.cap="Median Burden across Years and Household sizes"}
clean_years <- read_csv("../clean_data/clean_sample.csv")

@@

  This graph shows the development of median burden (proportion of housing costs relative to household income) from 1985 to 2013. The median burden varies dramatically between households of different sizes. The highest burden through the years has been on households of one person. This is understandable, since people living alone cannot split any costs between other earners.

@@

```{r, include = F, fig.cap = "Median Income Across Census Regions"}
url <- "http://www.mapsofworld.com/usa/usa-maps/united-states-regional-maps.html"
html <- read_html(url)
xpath <- '//table[@class="tableizer-table"]'

@@

  To gauge income levels in different areas we used the four census regions in the dataset and joined them with a scraped table from mapsoftheworld.com. Using the table we could connect census regions to state names and thus colour a map of the continental US with census region incomes. We used median incomes instead of mean incomes to avoid skewing the results with outliers.

  The household median monthly incomes vary between different census regions from 32200 dollars in the South to nearly 25% higher, 40000, in the West. The Northeast trails closely behind the West with 37267.50 dollars, and the Midwest is closer to the South with a median income of 35000.

@@

@@

@@

  The first plot compares APLmed by year for the four regions in the United States (Northeast, South, Midwest, West). APLmed is the adjusted median household income per person. This was precaculated using Housing and Urban Development (HUD) indexes. For example, a one person household's income is adjusted by a factor of 0.70, two by 0.80, and so on with a four person household being adjusted by 1.00 (or having no change from the median household income). The reason for no adjustment to a four person household income is because the median household size in the United States is about four persons.

@@

  The median adjusted income increases steadily (roughly linearly) over time. There is a decrease in income around 2010, correlated to the economic recession with a 2 year lag between the Northeast and the rest of the regions.The reason for this lag may be due to the fact that the financial center of the United States lies in the northeast, around New York City. So, when the 2008 crisis occurred the repercussions were immediately felt in the Northeast (Wall Street and Main Street- which refers to the Washington D.C. economy) while the other regions took some time to correct for the downturn in the financial market.

  The second plot of the graph depicts ABLmed versus year. ABLmed signifies the median adjusted income for the number of bedrooms in a household's living quarters. ABlmed was calculated using precaculated HUD indexes. Here, zero bedrooms was adjusted by a factor of 0.70, two bedrooms by 0.90, and so on. Three bedroom properties were adjusted by a factor of 1.04, because the median number of bedrooms in a property is roughly three bedrooms

@@

  This plot shows the amount of households in a given income group in a given year. Each household has been calculated a poverty treshold according to the Census Bureau’s official thresholds, based on family size and number of children.

@@

  As pictured in the graph, the relationship between household income and burden is seen as a positive correlation. From the graph, you can see that higher income household have less burden to deal with, which makes sense because those with more money are able to pay off monthly costs more easily. Those with 50% more burden are near the poverty line. From this graph, you can see a spike in households that have burden. Households with higher incomes started to pay more burdens, which can be attributed to 2007 financial housing market crisis.

@@

  From this graph, you can take a more indepth look at the relationship between housing costs and burden. It's interesting to see that all three burden lines are close to one another, which illustrates the principle that those with more money aren't buying more expensive houses. From the previous grpah, we saw that those with higher incomes have less burden, and those with lower incomes have higher burden. Each level of burden has similar housing costs, therefore reinforcing the idea that people are buying similar priced housing among the lower and middle classes. Again, a spike around 2010 is prevalent, which can be caused by the 2007 housing crisis.

@@

  From this graph, you can see that the Northeast and West have higher rates of burden at higher income levels, therefore illustrating how housing affordability in these regions are less than that of the Midwest and South. From this graph, you can see that the Northeast and West have higher rates of burden at higher income levels, therefore illustrating how housing affordability in these regions are less than that of the Midwest and South.

@@

  This plot charts a time series showing how the typical income for various types of households such as single family or apartment buildings changed over the years. One major trend is that typical income of all housing types except single family homes compared to their area has fallen, accelerating after 2007. However, incomes of single family homes, barring the housing crash in 2007, have been centered on the area's median income. There are a couple of possible explanations for this disparity. The first is that, with the growth of wages, more people could afford single family homes and thus leave the buildings with higher numbers of units. As wage grew, the area's median income also grew, and so, combined with higher waged workers moving into single family houses, single family homes remained constant while all other types shrunk proportionally. This would also explain the acceleration after the housing crisis, as single family homes became much cheaper temporarily and thus allow more workers to obtain single family housing. The second explanation is that as time went on, supply of housing increased, driving prices down and allowing lower waged laborers to obtain homes. Due to the purchasing power of higher wage laborers however, single family houses, even with their falling prices, were more difficult to obtain as time went on, and thus lower wage workers had to settle for other types of housing.

@@

 Further delving into the behavior of burden amongst the lower income workers, this plot shows how burden changes as a result of assistance based on what percentage of the area median income is their monthly housing costs. At initial observation, the burden of those who recieve more assistance is higher than those who recieve no assistance. This clearly makes sense, as those who need assistance are typically people who've lost their job or had some monetary crisis that would reduce their total salary. However, although assistance has around the same effect for most cost brackets, the households that have excessively high costs for their area have vastly higher burdens with assistance than all other brackets. One explanation for this is that those who purchase expensive houses in their area typically have higher salaries, and if they lose those salaries and require assistance they have to support the cost of the house. Also, single family homeowners have significantly less mobility so if they have a financial crisis they cannot easily transition to cheaper housing as opposing to those with rental properties typical of a larger number of units. The plot shows how burden changes as a result of assistance based on what percentage of the area median income is their monthly housing costs. At initial observation, the burden of those who recieve more assistance is higher than those who recieve no assistance. This clearly makes sense, as those who need assistance are typically people who've lost their job or had some monetary crisis that would reduce their total salary. However, although assistance has around the same effect for most cost brackets, the households that have excessively high costs for their area have vastly higher burdens with assistance than all other brackets. One explanation for this is that those who purchase expensive houses in their area typically have higher salaries, and if they lose those salaries and require assistance they have to support the cost of the house. Also, single family homeowners have significantly less mobility so if they have a financial crisis they cannot easily transition to cheaper housing as opposing to those with rental properties typical of a larger number of units.

>>>>

library(ggplot2)
library(dplyr)
library(readr)
library(stringr)
library(xml2)
library(rvest)
library(tidyr)
library(purrr)
library(scales)
library(maps)
library(ggthemes)
```
#Abstract

@@

  The dataset was taken from the American Housing Survey, which derives sets of data from 1985 and on. The data set includes categorizing each housing unit by factors such as affordability, income through fair market rent, burden, and the adjusted median income. Certain factors, such as Poverty Income, is based on the Census Bureau's official thresholds.

@@

@@

  The combined_years data was read into a dataframe using the read_csv function. Then, the combined years dataset was cleaned using the piping methods in dplyr. The data was rather messy, with unecessary quotations around character values and inconsistent letter cases in the column names. This was rectified by setting all the column names to lowercase and removing all the quotations from character values. One the main premises of this paper is comparing the burden of household that are owning and renting. Since there was no explicit labeling of data corresponding to owning and renting households, the "ownrent" column was mutated to a factor with the correct labels. Finally, the data was written to a clean_years csv file. The source files used different nonsensical integers (negative numbers) to indicate NA values. These were converted to NAs while reading the dataset in so that the output data and graphs made stastical sense - negative numbers where the values should be positive skew the dataset.

@@

  The main metric that is referred to in this paper is burden. Burden is defined by the proportion of household income that is spent towards the cost of housing. This metric is then further dissected by analyzing against various variables and collating data on contributing factors, such as income. The first data explored is income data across the four designated regions in the United States (Northeast, Midwest, South, West). The adjusted incomes (per persons and per bedroom) are also considered, as certain regions may have different averages for household and property sizes, which could skew the unadjusted income data. The behavior of household income, bracketed in ranges in accordance to census data, over time is displayed to provide a better understanding of the changes in burden over time. The household incomes are then compared versus burden over time. here it is found that the higher income levels generally have lower housing burdens, and those burdens actually decrased over time, while burdens in the lower income brackets were generally higher and increased over time. Also, the housing burdens in the Northeast and West are generally higher than the Midwest and South. Finally, to elucidate the gap in burden levels across high and low incomes, the housheold burden amongst owners and renters, and those with and without governement assistance is considered. Here it is found that owning households tend to have lower burden levels, as do those households with no assistance.

@@

```{r, echo = F, fig.cap="Median Burden across Years and Household sizes"}
clean_years <- read_csv("../clean_data/clean_years.csv")

@@

  Figure 1 shows the development of median burden (proportion of housing costs relative to household income) from 1985 to 2013. The median burden varies dramatically between households of different sizes. The highest burden through the years has been on households of one person. This is understandable, since people living alone cannot split any costs between other earners.

@@

```{r, echo = F, fig.cap = "Median Income Across Census Regions", message = F, warning=F}
url <- "http://www.mapsofworld.com/usa/usa-maps/united-states-regional-maps.html"
html <- read_html(url)
xpath <- '//table[@class="tableizer-table"]'

@@

  To map income levels in different areas we used the four census regions in the dataset and joined them with a scraped table from mapsoftheworld.com. Using the table we could connect census regions to state names and thus colour a map of the continental US with census region incomes. We used median incomes instead of mean incomes to avoid skewing the results with outliers, and filtered the map to include the most recent year, 2013.

  The resulting figure 2 shows household median monthly incomes vary between different census regions from $39,900 in the South to nearly 25% higher, $49,600, in the Northeast. The West trails closely behind the Northeast with $48,000, and the Midwest is closer to the South with a median income of $43,900.

@@

\newpage

@@

Figure 3: Median Adjusted Income per person (left) and per bedroom (right)

@@

  The first plot of figure 3 compares APLmed by year for the four regions in the United States (Northeast, South, Midwest, West). APLmed is the adjusted median household income per person. This was precaculated using Housing and Urban Development (HUD) indexes. For example, a one person household's income is adjusted by a factor of 0.70, two by 0.80, and so on with a four person household being adjusted by 1.00 (or having no change from the median household income). The reason for no adjustment to a four person household income is because the median household size in the United States is about four persons.

@@

  The median adjusted income increases steadily (roughly linearly) over time. There is a decrease in income around 2010, correlated to the economic recession with a 2 year lag between the Northeast and the rest of the regions.The reason for this lag may be due to the fact that the financial center of the United States lies in the northeast, around New York City. So, when the 2008 crisis occurred the repercussions were immediately felt in the Northeast (Wall Street) while the other regions took some time to correct for the downturn in the financial market.

  The second plot of figure 3 depicts the development of median income adjusted for the number of bedrooms in a household's living quarters. The adjusted median income was calculated using precaculated HUD indexes. Here, zero bedrooms was adjusted by a factor of 0.70, two bedrooms by 0.90, and so on. Three bedroom properties were adjusted by a factor of 1.04, because the median number of bedrooms in a property is roughly three bedrooms

@@

  Figure 4 shows the amount of households in a given income group in a given year. Each household has been calculated a poverty treshold according to the Census Bureau’s official thresholds, based on family size and number of children.

@@

  As pictured in figure 5, the relationship between household income and burden is seen as a positive correlation. From the graph, you can see that higher income household have less burden to deal with, which makes sense because those with more money are able to pay off monthly costs more easily. Those with 50% more burden are near the poverty line. Figure 5 also shows a spike in households that have burden. Households with higher incomes started to pay more burdens, which can be attributed to 2007 financial housing market crisis.

@@

  Figure 6 offers a more in-depth look at the relationship between housing costs and burden. It's interesting to see that all three burden lines are close to one another, which illustrates the principle that those with more money aren't buying more expensive houses. From figure 5 we saw that those with higher incomes have less burden, and those with lower incomes have higher burden. Each level of burden has similar housing costs, therefore reinforcing the idea that people are buying similar priced housing among the lower and middle classes. Again, a spike around 2010 is prevalent, which can be caused by the 2007 housing crisis.

@@

  From figure 7 you can see that the Northeast and West have higher rates of burden at higher income levels, therefore illustrating how housing affordability in these regions are less than that of the Midwest and South.

@@

  Figure 8 charts a time series showing how the typical income for various types of households such as single family or apartment buildings changed over the years. One major trend is that typical income of all housing types except single family homes compared to their area has fallen, accelerating after 2007. However, incomes of single family homes, barring the housing crash in 2007, have been centered on the area's median income. There are a couple of possible explanations for this disparity. The first is that, with the growth of wages, more people could afford single family homes and thus leave the buildings with higher numbers of units. As wages increased, the area's median income also grew, and so, combined with higher waged workers moving into single family houses, single family homes remained constant while all other types shrunk proportionally. This would also explain the acceleration after the housing crisis, as single family homes became much cheaper temporarily and thus allow more workers to obtain single family housing. The second explanation is that as time went on, supply of housing increased, driving prices down and allowing lower waged laborers to obtain homes. Due to the purchasing power of higher wage laborers however, single family houses, even with their falling prices, were more difficult to obtain as time went on, and thus lower wage workers had to settle for other types of housing.

@@

 Further delving into the behavior of burden amongst the lower income workers, figure 9 shows how burden changes as a result of assistance based on what percentage of the area median income is their monthly housing costs. At initial observation, the burden of those who receive more assistance is higher than those who receive no assistance. This clearly makes sense, as those who need assistance are typically people who've lost their job or had some monetary crisis that would reduce their total salary. However, although assistance has around the same effect for most cost brackets, the households that have excessively high costs for their area have vastly higher burdens with assistance than all other brackets. One explanation for this is that those who purchase expensive houses in their area typically have higher salaries, and if they lose those salaries and require assistance they have to support the cost of the house. Also, single family homeowners have significantly less mobility so if they have a financial crisis they cannot easily transition to cheaper housing as opposing to those with rental properties typical of a larger number of units. The plot shows how burden changes as a result of assistance based on what percentage of the area median income is their monthly housing costs.

 At initial observation, the burden of those who recieve more assistance is higher than those who recieve no assistance. This clearly makes sense, as those who need assistance are typically people who've lost their job or had some monetary crisis that would reduce their total salary. However, although assistance has around the same effect for most cost brackets, the households that have excessively high costs for their area have vastly higher burdens with assistance than all other brackets. One explanation for this is that those who purchase expensive houses in their area typically have higher salaries, and if they lose those salaries and require assistance they have to support the cost of the house. Also, single family homeowners have significantly less mobility so if they have a financial crisis they cannot easily transition to cheaper housing as opposing to those with rental properties typical of a larger number of units.
